---
title: 'Running `R` on server'
author: 'Pratik Bhandari'
institute: 'PLEAD, VD-mit'
date: '10/06/2022'
toc: false
format:
  beamer:
    incremental: false
    slide-level: 2
    theme: CambridgeUS
    colortheme: dolphin
    highlight: zenburn
    aspectratio: 169
---

## Copy files from local machine to your coli home folder

```{r, fileCopy1}
#| eval: false
#| echo: true
scp -r ./analyses username@login.coli.uni-saarland.de:/home/CE/username/expt

```

Copying from the `analyses` folder in the local working directory to `expt` in the `login` home.

A password prompt will appear. Provide the password that you use to login to your coli email.

## Create a secure connection

Connect to the coli login

```{r, coliConnect}
#| eval: false
#| echo: true

ssh username@login.coli.uni-saarland.de
```

Again, provide your coli login credentials as a prompt appears.

## Copy files from coli login to desired server

`jones-3` is just an example here.

```{r, fileCopy2}
#| eval: false
#| echo: true

scp -r username@login.coli.uni-saarland.de:/home/CE/username/expt \
username@jones-3:/home/CE/username/expt/analysis
```

Copying from `expt` in the `coli login` to `analysis` in `jones-3` server.

## Connect to the server

```{r, jonesConnect}
#| eval: false
#| echo: true
# From the coli login:

ssh -X username@jones-3.coli.uni-saarland.de

#Or:

ssh jones-3
```

## Navigate to your home directory and open R

Change my current working directory from `/home/CE/username` to `./expt/analysis`.\
And open `R` in this directory.

```{r, cd}
#| eval: false
#| echo: true
pwd 
cd expt/analysis

R
```


## Find R libraries

```{r, findLib}
#| eval: false
#| echo: true
.libPaths()

```

Likely output is:

```{r, showLib}
#| eval: false
#| echo: true
[1] "/usr/local/lib/R/site-library"                           
[2] "/usr/lib/R/site-library"                                 
[3] "/usr/lib/R/library"  
```

## Add R libraries

```{r, addLib}
#| eval: false
#| echo: true
.libPaths(new="/local/home/CE/pratikb/R/x86_64-pc-linux-gnu-library/4.0") 
```

By this, you are adding additional path to R libraries which points to the path of `pratikb`'s libraries so that you don't have to install packages required for LMM again.  

**Warning**: I haven't tried this, so can't tell if it gives any error.

*Resources*:

- 1: Link -- [stackoverflow-1](https://stackoverflow.com/questions/31707941/how-do-i-change-the-default-library-path-for-r-packages#31707983)

- 2: Link -- [stackoverflow-2](http://stackoverflow.com/questions/40848776/ddg#40849165)

- 3: Link -- [stackoverflow-3](https://stackoverflow.com/questions/15170399/change-r-default-library-path-using-libpaths-in-rprofile-site-fails-to-work)

- 4: Check documentation `help(.libPaths)`

## Load prerequsities for analyses

<!-- Now that you have your libraries set up, and data files are already imported, you can start running analyses in `R` on the server. -->
Load packages, data files, functions, etc. and run your analyses.

```{r, load}
#| eval: false
#| echo: true

library(lme4)
library(dplyr)

load('analyses/datfiles/dat.RData')
source('analyses/functions/customfn.R')

```

**N.B.**: It is clear that the `analyses` folder that was imported from your computer to the server should have contained datafiles and functions.

## Run the analyses of interest

You know what to do next -- just run the good old lmer model.\

```{r, runLmer}
#| eval: false
#| echo: true

m0 <- glmer(...)

```

Save the model output.`\
Export it as an RDS file (e.g. `results.RDS`), or export multiple model outputs as one .RData file.\
Then copy them back to your computer.\

## Shortcut

It is not always necessary to hand-run the models in R.\
You can write a single script with `library`, `load`, `source`, `lmer`, `save`, etc. and run it in the background.\

`One potential drawback`: You might not be able to see what/where the error occurs.\
But you can run them in small chunks anyway. Or, maybe create a custom alert message? I don't have an experience with it; it should be possible, in principle.

## Copy back

First copy the outputs to the coli login then to the local machine.

1. Run this in the jones-3 terminal:

```{r, jonestocoli}
#| eval: false
#| echo: true
#| results: asis

scp -r  username@jones-3.coli.uni-saarland.de:/home/CE/username/
  expt/analyses/results.RDS \
username@login.coli.uni-saarland.de:/home/CE/username/expt

```

This copies the files from the jones server to the coli login.

## Copy back

2. Run this in your local terminal\

```{r colitolocal}
#| eval: false
#| echo: true
scp -r username@login.coli.uni-saarland.de:/home/CE/username/expt ./analyses/results

```

This copies the files from `expt` in coli login to the local machine.

## Final thoughts

- Everything discussed today can be further simplified with some (?smart?) bash scripting.\

- It might not always be useful to run our (?my?) analyses on the server.

    - Time-effort and long-term usefulness tradeoff :wink:

- Familiarizing oneself with basic bash commands is **very** helpful regardless of running R on server.

- It is fun to try new things anyway :smile:

**Let me know if there is anything wrong and/or to correct or improvise here.**

## Resourceful people

Please consult:

- [Iza](https://www.uni-saarland.de/lehrstuhl/demberg/members/iza-skrjanec.html) and [Margarita](https://www.uni-saarland.de/lehrstuhl/demberg/members/rita.html) for general approach about connecting to server and R in particular respectively.

- [Marian](https://www.uni-saarland.de/lehrstuhl/demberg/members/marian.html) about GUI approach to connect to the server.


